---
title: "Scenario 3: R Markdown to PDF"
author: "Yi Zhang"
date: "2026-08-31"
output:
  pdf_document:
    toc: true
    number_sections: true
---

<!--
============================================================
场景 3：.Rmd —— 551 交 lab 走的就是这条路，最后要出 PDF。

写作时按块跑：
  ⇧⌘Enter   跑光标所在的代码块
  ⌘⌥P       跑当前块以上的所有块（重开文件后恢复上下文）
  ⇧⌘V       预览

出 PDF：
  ⇧⌘K                        Knit（按 YAML 里的 output 走，这里就是 PDF）
  ⇧⌘P → "R: Knit Rmd To PDF"  强制出 PDF

⚠️ 标题、正文、图里都不能有中文——pdflatex 会直接报
   "Unicode character not set up for use with LaTeX"。
   所以这个文件正文全是英文，注释写在这种 HTML 注释里（不进 PDF）。
============================================================
-->

```{r setup, include=FALSE}
# include=FALSE：这个块会执行，但代码和输出都不进 PDF。
# 每个 .Rmd 开头基本都有这么一块，用来设全局选项。
knitr::opts_chunk$set(
  echo = TRUE,        # 默认把代码也印进 PDF
  warning = FALSE,    # 不印 warning
  message = FALSE,    # 不印 library() 那堆加载信息
  fig.width = 6,
  fig.height = 3.5
)
library(tidyverse)
```

# What this file is

This is the workflow DSCI 551 labs use: write prose and R code in one
`.Rmd` file, then render straight to PDF.

Three things to notice: chunk options control what reaches the PDF,
inline code lets you put computed numbers inside sentences, and the
whole document runs top-to-bottom in a **fresh** R session when knitted.

# Chunk options

<!-- 下面演示三种最常用的 chunk 选项 -->

The default: code shown, output shown.

```{r}
summary(mtcars$mpg)
```

`echo=FALSE` hides the code but keeps the output. Lab questions that ask
for a plot usually want this.

```{r echo=FALSE}
mtcars |>
  count(cyl) |>
  rename(cylinders = cyl, cars = n)
```

`eval=FALSE` shows the code but does not run it.

```{r eval=FALSE}
# This line is never executed.
install.packages("something")
```

# Inline code

The `mtcars` dataset has `r nrow(mtcars)` rows and `r ncol(mtcars)`
columns. Mean fuel economy is `r round(mean(mtcars$mpg), 2)` mpg.

<!--
上面那三个反引号包起来的行内代码，渲染后会变成实际数字。
写结论句子的时候用这个，比手敲数字安全——数据变了句子自动跟着变。

⚠️ 坑：knitr 连 HTML 注释里的行内代码也会执行。
   所以注释里不能原样写那个语法，否则 knit 会报
   "'...' used in an incorrect context" 然后整个渲染失败。
   （这个坑我踩了，就是在这一行。）
-->

# A figure

```{r fig-scatter, fig.cap="Weight versus fuel economy, coloured by cylinder count."}
ggplot(mtcars, aes(x = wt, y = mpg, colour = factor(cyl))) +
  geom_point(size = 2.5) +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon", colour = "Cylinders") +
  theme_minimal()
```

<!--
交叉引用（\ref{...}）在普通 pdf_document 下解析不出来，会印成 "??"，
要用 bookdown::pdf_document2 才行。lab 里一般用不上，先不引入。
-->

# A table

```{r}
knitr::kable(
  head(mtcars[order(-mtcars$mpg), c("mpg", "cyl", "wt", "hp")], 5),
  caption = "Five most fuel-efficient cars."
)
```

# Math

LaTeX math works because the PDF goes through LaTeX. DSCI 551 will need
this constantly.

Inline: the mean is $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$.

Display:

$$
\hat{\theta}_{\text{MLE}} = \arg\max_{\theta} \prod_{i=1}^{n} f(x_i \mid \theta)
$$

# The one trap

<!--
这一段是真正要记住的东西。
-->

Knitting runs the document in a **brand new R session**. Anything you
created interactively in your terminal does not exist there.

So a document can work perfectly while you click through it, and still
fail to knit. Before submitting, always knit once from scratch and check
the PDF actually came out.

```{r}
# This works when knitted because tidyverse is loaded in the setup chunk
# above. If it were only loaded in your terminal, knitting would fail.
mtcars |> summarise(n = n(), mean_mpg = mean(mpg))
```
